The Talking Heads experiment

### 2015-04-29 chap12/chap12.tex
corrected some typos in Ch. 10-12

But the skeptical observer will argue that most of these experiments are about language emergence rather than 
language evolution per se. Agents are given a certain set of strategies, which thus constitutes their 
innate Language Acquisition Device, and it is then demonstrated by computer simulations or 
robotic experiments (i) how these strategies are effective to {\itshape acquire} an existing language system in a 
population, (ii) how they are effective to {\itshape extend} a language system so that it remains adaptive 
to the needs and environments of its user, and (iii) how a population can {\itshape invent and coordinate} a new language system from 
scratch by using the strategy.

Although long-term change is possible
given a particular language strategy (for example, new color categories and color terms might emerge
when needed to cope with new situations in the environment), this change does not alter
the set of language strategies available to the agent. Therefore - the critical observer would say - all this does still
not address the question yet where the strategies come from.

@@

To clarify these issues, let us first define a distinction between emergence, language change, and evolution:\is{culture-driven language
evolution}
\begin{itemize}
\item {\itshape Language emergence} is the phenomenon whereby a language system, i.e. a set of conventions and conceptualisations
for a particular semantic domain, for example tense-aspect, color, space or determination, arises in a population and 
becomes shared, based on a particular strategy (or a set of strategies) that is shared by all agents. 
\item {\itshape Language change} happens when the components of a language system and its conceptualisations shift, but the 
change remains within the bounds of a particular strategy. For example, the boundaries how different cases are used 
may shift (as is happening with the dative and genitive at the moment in German) or a new basic color category and
color word may arise and claim a region within the color space (as happened with the word `orange' which 
started to be used as a basic color word in 16th century Middle English to occupy a distinctive region between yellow and red). 
\item {\itshape Language evolution} occurs when new strategies arise. There is plenty of evidence from creole languages and 
from historical linguistics that this happens at a cultural time scale. Here are three examples:
\begin{enumerate}
\item Phrase structure (in the sense of 
strict word order based on abstract hierarchical patterns) emerged gradually in 
proto-Indo European, from a stage with largely free word order but a strong inflectional system \citep{VandeVelde:2011}. 
\item Word order and prepositions became the dominant strategy instead of morphological cases 
for expressing argument structure in the late Old English period \citep{VanKemenade:1987}. 
\item There is currently an evolution going on in Spanish clitics (`le', `la', `lo') whereby the etymological system 
of Standard Spanish, which uses clitics to express different cases (nominative, dative, accusative), is shifting 
to a referential system (expressing gender and number). Case differentiation is lost, but existing forms are recruited
for the new functions implied by a referential strategy  \citep{Fernandez-Ordonez:1999}. 
\end{enumerate}
\end{itemize}
It is not always obvious from the observation of language facts whether we are dealing with change (within the boundaries 
of a strategy) or with a new strategy. However, this becomes entirely clear when we try to computationally model the processes of 
acquisition, expansion or invention of the language strategy involved. We always try to understand change first using existing 
strategies until we are forced to introduce a new strategy because the differences at the conceptual or linguistic 
level are too profound to computationally handle them.

The distinctions made here map to similar ones in biology. Biological strategies encoded in the genome govern the 
construction of phenotypes which include the form of organisms and their behaviors 
ensuring survival issues such as predator avoidance, food gathering, navigation, mate finding, nest building, etc. 
When we see a swarm of birds in the sky, or a path formed by ants, or a beaver dam, we see an example of 
{\itshape emergence}: A large-scale macroscopic structure emerges from the individual activities of the agents (i.e. animals)
without central control, similar to the way a language system emerges as a macroscopic shared structure
through the individual activities of the linguistic agents. The activities of the ants or the beavers are all governed by the 
same innate strategy. For example, to achieve path formation, ants put down pheromone when returning from the nest
and they follow pheromone gradients to find a path.

Living organisms are {\itshape adaptive} to their environment and so even within 
the bounds of a given set of innate strategies there may still be significant changes observed in appearance, 
behavior or internal functioning. For example, beavers 
may start using rocks for dams when mud and branches are not available, birds use plastics or other waste products, 
a goat may start to walk on two hind legs when it did not grow the forelegs \citep{Jablonka:2013}. 
Nevertheless true {\itshape evolution} is only said to happen if there are new innate 
strategies, i.e. genetically coded strategies, that arise, for example if a fish species genetically 
peciates into subspecies where one continues to bread eggs deposited on the bottom of the sea whereas the 
new varient does mouthbrooding, meaning that eggs are bred within the mouth of a parent. 
Also in biology it is not always clear whether we are dealing with adaptive change (without genomic change) or evolution
(with genomic change). Some variants that were thought to be a different species at one point of time
turned out to be ecotypes of the same species, i.e. variants that 
have the same genetics but a different appearance or behavior due to adaptation to the environment. However today these debates
can be entirely settled when a genetic analysis is performed.

For biological organisms, the standard approach to find out how (innate) strategies evolved is through the framework  
of genetic evolution by natural selection and it is logically possible that language strategies are also innate, in which 
case we should be able to figure out how they are genetically coded and evolved. 
There are indeed several researchers who strongly favor this hypothesis and they group their work 
under the banner of biolinguistics \citep{Boeckx:2013}.

But it is also possible that language strategies are the outcome of cognitive development and cultural evolution, 
just like humans have developed a multitude of non-innate strategies in a wide range of domains: strategies for finding your 
way in cities, for building shelters, for preparing food, for playing Jazz piano, for fixing technical equipment, 
for organising trade, composing music, negotiating a contract, finding a partner, etc.  
This is the idea behind the {\itshape recruitment theory} of language evolution, which postulates that cognition and 
cultural evolution are the primary motors for the origins and propagation of new language strategies, even though of 
course they require the necessary neurobiological capacities that make each strategy possible. Once a strategy culturally 
evolved, it exerts a pull effect on genetic evolution, because genetic variants are favored that can carry out a strategy more easily. 
So this hypothesis implies {\itshape culture-driven gene evolution} rather than gene-driven cultural 
evolution \citep{Fisher:2013}.

@@

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical contruction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, since  a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

\begin{itemize}
\item {\itshape Expressive adequacy}: The language systems available to the speaker and the hearer must have 
the necessary expressive adequacy to reach communicative success. 
The available conceptual systems must include the needed conceptual distinctions 
and the available linguistic systems must be able to express these distinctions. Expressive adequacy depends 
not only on the goals of the speaker and the available embodiment but also on the environment. For example, 
if only black-and-white pictures are available, a strategy based on hue is entirely inadequate. On the other hand, 
if one of the dialog partners is color blind, a hue-based strategy would not work for colored pictures. 
\is{expressive adequacy}

@@

Organisms use a variety of strategies for survival and offspring production:
bodily shape and appearance, specific sensors and actuators, metabolism, 
innate behaviors. Biologists call the ecological area that a particular organism occupies 
thanks to its physiological and behavioral strategies its ecological {\itshape niche}. A niche is defined through the ecological conditions
(temperature, altitude, habitat type, food availability, presence of predators, etc.) for which the organism's strategies
are appropriate and hence within which the species reaches a sufficiently high fitness to survive and fend off predators. 
Organisms may also have strategies to alter their environment so that some of their strategies become more possible or more optimal, 
a process called {\itshape niche construction} \citep{Odling-Smee:2003} \is{fitness landscape}

@@

The previous chapter already showed that there are 
different strategies for color language: basic color terms, graded membership (`very green'), compounds (`sky blue'), 
brightness qualification (`light green'), etc. All these strategies use a particular color space, more specifically 
the CIE L*u*v* space where the L* dimension represents brightness (ranging from black to white), the u* 
dimension represents the red-green opponent channel, 
and the v* dimension the yellow-blue channel. This space was chosen 
because the distance between two colours in the L*u*v* space accurately represents 
the psychological distances between these colours as perceived by human subjects. \is{Color Description Game}

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): \enlargethispage{2\baselineskip}
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-color strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

\begin{figure}[htb!]
\centerline{
\includegraphics[width=0.6\textwidth]{chap12/figs/selective-advantage-bar.pdf}
}
\caption{{Niche analysis for three different 
color strategies: brightness-only, brightness-hue, and hue-only. 
The properties of the context vary on the x-axis and the y-axis shows performance of each strategy.
Only one niche dimension is examined, namely communicative success. 
We see that each strategy has its own niche for which it is optimal, but the full-color strategy is globally 
performing the best across contexts, which may explain why it has become preferred in most language families. 
\label{fig:bar}}
} 
\end{figure}

Bleys initialised the three populations with the three different strategies and let them first learn an interpretation of 
the Spanish focal colors (with prototypes in brightness\-only, hue-only, and full-colour spaces). Then he measured 
performance for communicative success for the varying ecological conditions. 
Not unexpectedly, he found that depending on the percentage of the importance of hue differences in the context, a different strategy 
is more optimal (see \figref{fig:bar}). Not surprisingly, a hue-only strategy fails completely when the samples in each  
context have equal hue but different brightness, whereas a brightness-only strategy fails when the samples have 
equal brightness but different hue.

@@

The Bleys experiment concerns the brightness-based and the full-colour strategy. Both use 
a prototype-based nearest neighbor categorisation. We know from earlier experiments with the Proper Naming Game
and the Action Naming Game that this strategy enables a population of agents 
to self-organise a colour lexicon from scratch. \figref{fig:full-cs-dynamics} illustrates (on the left) how 
communicative success increases and the lexicon becomes more coherent, and (on the right) how the prototypes
of the different agents gradually expand and become similar to be almost identical after 1200 games per agent, when 
the system has stabilised to 6 basic colors.

@@

The {\itshape Brightness Prototype Strategy} is similar to the Basic Color Strategy, but instead of taking all three dimensions into account, only the L* dimension of both the stimuli in the context and the prototypes of the colour categories are compared. While learning, the prototype of the used colour category is shifted on the L* axis towards the L* value of the topic. During invention, only the L* value of the topic is considered relevant. \figref{fig:brightness-dynamics} shows that this strategy is also adequate to allow a population of agents to self-organise and coordinate a colour lexicon from scratch. The resulting colour lexicon now consists of different shades of gray. The set of prototypes also expands and becomes similar across agents.

@@

Agents are initialised with a random value for the strategy-score for the Brightness-based versus
Full-colour based strategy. When the dynamics described here takes its 
course (which includes the shifting of prototypes by speakers and hearers to be maximally adapted to the 
contexts that are presented), we observe several possibilities. In one simulation run (see \figref{fig:strategies-one}), 
the brightness-based strategy becomes and remains dominant. It could have been the brightness 
or the full-colour strategy, depending on small fluctuations in the early stages. 
In another simulation run we see that one strategy becomes dominant first (in this case the brightness strategy) 
to be overtaken later by another strategy (i.c.\ the full-colour strategy), 
as attested in the history of English color terms (see \figref{fig:strategies-dynamics}). 
The two strategies continue to co-exist. Brightness is still used in 
circumstances where this gives a higher chance of communicative success, for example when colour chips are 
close in hue but distinct in brightness or when there is a word which has most of its success in the brightness 
dimension.

@@

\begin{figure}
\centerline{\includegraphics[width=.9\linewidth]{chap11/figs/near-the-box2.pdf}}
\caption{Example of a conceptualisation for the phrase `near to the box'. {\itshape Top:} 
Leftmost: The situation as perceived by the robot. Middle: situation after a geometric transform where the box is 
made the central perspective. Right: Area around the box which reflects (in terms of darkness of the color) how 
near objects are to the box. {\itshape Bottom:} IRL constraint network for this conceptualisation.
\label{fig:spatial-strategies}}
\end{figure}

@@

Once an adequate conceptualisation is found and it leads to a successful game, modular sections of the network can be 
abstracted and used as the basis of reusable conceptualisation strategies in the future. Each conceptualisation 
strategy has its own score which is updated using the lateral inhibition dynamics as used for 
the lexical or grammatical contructions as explained in earlier chapters. The frequency of use and success 
of a strategy depends partly on whether the ``niche'' in which a strategy is most effective occurs frequently, partly 
on what is preferred by the rest of the population, and other criteria such as complexity.

@@

Are there general meta-strategies that can be used to generate the kind of lexico\-gram\-mat\-ical strategies 
that can in turn generate all the language systems we observe 
in today's languages? This is a very profound question, related to the ongoing debate in linguistics on the existence of 
a universal language acquisition device. A lot of ink has been spilled but we need to do profound computational modeling 
of meta-strategies and thorough experiments to find out.

@@

For {\itshape lexical strategies}, the diagnostics consist in checking whether there is part 
of the meaning that is not yet covered by any lexical item. The repair then consists in introducing 
a (possibly new) word and introducing a new lexical construction for this word. 
This strategy is already used by the Talking Heads. When part of a meaning cannot be covered, a new wordform is randomly generated
and used from then on for this meaning, but possibly in competition with other words that were later acquired for the same 
meaning.

\newpage
A study of ongoing change in the vocabularies of human speakers shows that the invention of a brand new wordform happens seldom. 
Usually an existing wordform is recruited to express an uncovered meaning. The meaning of the already existing wordform
expresses aspects of the novel meaning to be expressed but there is not an exact overlap. However, the existing word is sufficiently 
suggestive so that the hearer can guess what meaning is intended by the speaker. For example, the word `mouse' becomes 
adopted for computer mouse based on its appearance. Often a compound of several words is used, each of them suggesting 
other aspects of the meaning, such as `bitcoin', which suggests digital (`bit') and currency (`coin').  
There are several other strategies, for example, an abbreviation of a long word becomes itself a single new word, 
such as `MOOC', meaning `massively open on-line course' and now standing for a course made available through the internet. 
There are also playful variations created with existing words, as for example
`wackadoo' which means `bizarre person' and probably originally came from the word `wacky'.

@@

Historical linguists have researched many concrete examples of how agreement systems arise.\footnote{See for example: \cite{Lehmann:1988}, \cite{Luraghi:2011}.}
There are invariably the following phases: 
\begin{enumerate}
\item Agreement markers derive from reusing existing words, such as pronouns or classifiers. 
\item The markers derived from independent words become shorter, they lose part of their form, 
then become clitics and later affixes. For example, the Dyirbal (Australian aboriginal)
agreement marker {\itshape m-} evolved from the classifier {\itshape mayi}, which means 'non-flesh food'. 
\item The features that agreement markers express, which are initially semantically 
grounded, invariably become more abstract and get to be used in 
a purely conventional manner. For example, the masculine/feminine 
gender distinction has its basis in male/female sex, but is then arbitrarily applied to inanimate objects, so that table 
might be masculine in one language (German: {\itshape der Tisch}) but feminine in another (French: {\itshape la table}). 
\end{enumerate}
Starting from a game of reference, similar to the Talking Heads experiment, populations of agents were programmed 
first to build a ``formal'' agreement system, i.e. a system based on formal markers without any meaning, and then 
a meaningful grammatical system, with grammatical markers that express semantic features. The experiment was 
then extended with further strategies that optimise three niche factors:
\begin{enumerate}
\item The recruitment of existing words as markers makes the acquisition of new markers more accurate because it is based 
on existing words and so the meaning can be guessed more easily. This helps to propagate the markers faster in the population. 
\item The phonological reduction of wordforms leads to less articulatory effort, as we have already seen with the German 
article system. 
\item Conventionalisation leads to a smaller inventory, and hence less memory requirements and better preservation of the 
language system. 
\end{enumerate}

>>>>

But the sceptical observer will argue that most of these experiments are about language emergence rather than 
language evolution per se. Agents are given a certain set of strategies, which thus constitutes their 
innate Language Acquisition Device, and it is then demonstrated by computer simulations or 
robotic experiments (i) how these strategies are effective to {\itshape acquire} an existing language system in a 
population, (ii) how they are effective to {\itshape extend} a language system so that it remains adaptive 
to the needs and environments of its user, and (iii) how a population can {\itshape invent and coordinate} a new language system from 
scratch by using the strategy.

Although long-term change is possible
given a particular language strategy (for example, new colour categories and colour terms might emerge
when needed to cope with new situations in the environment), this change does not alter
the set of language strategies available to the agent. Therefore - the critical observer would say - all this does still
not address the question yet where the strategies come from.

@@

To clarify these issues, let us first define a distinction between emergence, language change, and evolution:\is{culture-driven language
evolution}
\begin{itemize}
\item {\itshape Language emergence} is the phenomenon whereby a language system, i.e. a set of conventions and conceptualisations
for a particular semantic domain, for example tense-aspect, colour, space or determination, arises in a population and 
becomes shared, based on a particular strategy (or a set of strategies) that is shared by all agents. 
\item {\itshape Language change} happens when the components of a language system and its conceptualisations shift, but the 
change remains within the bounds of a particular strategy. For example, the boundaries how different cases are used 
may shift (as is happening with the dative and genitive at the moment in German) or a new basic colour category and
colour word may arise and claim a region within the colour space (as happened with the word `orange' which 
started to be used as a basic colour word in 16th century Middle English to occupy a distinctive region between yellow and red). 
\item {\itshape Language evolution} occurs when new strategies arise. There is plenty of evidence from creole languages and 
from historical linguistics that this happens at a cultural time scale. Here are three examples:
\begin{enumerate}
\item Phrase structure (in the sense of 
strict word order based on abstract hierarchical patterns) emerged gradually in 
proto-Indo European, from a stage with largely free word order but a strong inflectional system \citep{VandeVelde:2011}. 
\item Word order and prepositions became the dominant strategy instead of morphological cases 
for expressing argument structure in the late Old English period \citep{VanKemenade:1987}. 
\item There is currently an evolution going on in Spanish clitics (`le', `la', `lo') whereby the etymological system 
of Standard Spanish, which uses clitics to express different cases (nominative, dative, accusative), is shifting 
to a referential system (expressing gender and number). Case differentiation is lost, but existing forms are recruited
for the new functions implied by a referential strategy  \citep{Fernandez-Ordonez:1999}. 
\end{enumerate}
\end{itemize}
It is not always obvious from the observation of language facts whether we are dealing with change (within the boundaries 
of a strategy) or with a new strategy. However, this becomes entirely clear when we try to computationally model the processes of 
acquisition, expansion or invention of the language strategy involved. We always try to understand change first using existing 
strategies until we are forced to introduce a new strategy because the differences at the conceptual or linguistic 
level are too profound to computationally handle them.

The distinctions made here map to similar ones in biology. Biological strategies encoded in the genome govern the 
construction of phenotypes which include the form of organisms and their behaviours 
ensuring survival issues such as predator avoidance, food gathering, navigation, mate finding, nest building, etc. 
When we see a swarm of birds in the sky, or a path formed by ants, or a beaver dam, we see an example of 
{\itshape emergence}: A large-scale macroscopic structure emerges from the individual activities of the agents (i.e. animals)
without central control, similar to the way a language system emerges as a macroscopic shared structure
through the individual activities of the linguistic agents. The activities of the ants or the beavers are all governed by the 
same innate strategy. For example, to achieve path formation, ants put down pheromone when returning from the nest
and they follow pheromone gradients to find a path.

Living organisms are {\itshape adaptive} to their environment and so even within 
the bounds of a given set of innate strategies there may still be significant changes observed in appearance, 
behaviour or internal functioning. For example, beavers 
may start using rocks for dams when mud and branches are not available, birds use plastics or other waste products, 
a goat may start to walk on two hind legs when it did not grow the forelegs \citep{Jablonka:2013}. 
Nevertheless true {\itshape evolution} is only said to happen if there are new innate 
strategies, i.e. genetically coded strategies, that arise, for example if a fish species genetically 
speciates into subspecies where one continues to bread eggs deposited on the bottom of the sea whereas the 
new variant does mouthbrooding, meaning that eggs are bred within the mouth of a parent. 
Also in biology it is not always clear whether we are dealing with adaptive change (without genomic change) or evolution
(with genomic change). Some variants that were thought to be a different species at one point of time
turned out to be ecotypes of the same species, i.e. variants that 
have the same genetics but a different appearance or behaviour due to adaptation to the environment. However today these debates
can be entirely settled when a genetic analysis is performed.

For biological organisms, the standard approach to find out how (innate) strategies evolved is through the framework  
of genetic evolution by natural selection and it is logically possible that language strategies are also innate, in which 
case we should be able to figure out how they are genetically coded and evolved. 
There are indeed several researchers who strongly favour this hypothesis and they group their work 
under the banner of biolinguistics \citep{Boeckx:2013}.

But it is also possible that language strategies are the outcome of cognitive development and cultural evolution, 
just like humans have developed a multitude of non-innate strategies in a wide range of domains: strategies for finding your 
way in cities, for building shelters, for preparing food, for playing Jazz piano, for fixing technical equipment, 
for organising trade, composing music, negotiating a contract, finding a partner, etc.  
This is the idea behind the {\itshape recruitment theory} of language evolution, which postulates that cognition and 
cultural evolution are the primary motors for the origins and propagation of new language strategies, even though of 
course they require the necessary neurobiological capacities that make each strategy possible. Once a strategy culturally 
evolved, it exerts a pull effect on genetic evolution, because genetic variants are favoured that can carry out a strategy more easily. 
So this hypothesis implies {\itshape culture-driven gene evolution} rather than gene-driven cultural 
evolution \citep{Fisher:2013}.

@@

Today, Darwinian evolutionary dynamics \is{Darwinian evolutionary dynamics}
is recognised as a general principle that can and has been applied to 
many different types of systems, in chemistry, neurobiology or 
economics.\footnote{See \cite{Arthur:1996}, \cite{Edelman:1987} and \cite{Luisi:2001}.}
It has also been used successfully to come up 
with new artificial systems, for example to evolve controllers for robots using 
genetic algorithms \citep{Nolfi:2000}. It is therefore not so surprising that 
selectionism can also be mapped to the cultural and more specifically to the linguistic domain in the following way: 
\begin{enumerate}
\item Speakers generate variants, either accidentally through errors in their speech production, or deliberately, 
for example, when they try to express a novel idea by inventing new forms or 
appropriating existing forms for new purposes, for example, by coercing a word or grammatical construction into a new usage. 
\item Which variants will survive in the communal language depends on selectionist criteria. I will 
discuss them shortly in more detail. They include for example whether the speaker is understood by the hearer, or whether 
it reduces the cognitive complexity of parsing and producing and hence allows larger sentences to be understood. 
\item The self-enforcing causal loop happens when variants created by one speaker start to propagate in the population 
and begin to replace their competitors. For example, a particular word to express a given meaning gets out of fashion 
and is taken over by another one. Some of these developments are simply fashion trends, but the more profound ones, 
in particular the ones that involve grammar, are related to critical selectionist criteria. 
\end{enumerate}
This Darwinian dynamic is already active in the operation of language strategies, since  a strategy regulates the 
creation and selection of variants, determines how feedback has to be handled, and implies a self-enforcing causal loop through 
lateral inhibition so that variants satisfying the selection criteria become 
dominant in the population.

@@

\begin{itemize}
\item {\itshape Expressive adequacy}: The language systems available to the speaker and the hearer must have 
the necessary expressive adequacy to reach communicative success. 
The available conceptual systems must include the needed conceptual distinctions 
and the available linguistic systems must be able to express these distinctions. Expressive adequacy depends 
not only on the goals of the speaker and the available embodiment but also on the environment. For example, 
if only black-and-white pictures are available, a strategy based on hue is entirely inadequate. On the other hand, 
if one of the dialogue partners is colour blind, a hue-based strategy would not work for coloured pictures. 
\is{expressive adequacy}

@@

Organisms use a variety of strategies for survival and offspring production:
bodily shape and appearance, specific sensors and actuators, metabolism, 
innate behaviours. Biologists call the ecological area that a particular organism occupies 
thanks to its physiological and behavioural strategies its ecological {\itshape niche}. A niche is defined through the ecological conditions
(temperature, altitude, habitat type, food availability, presence of predators, etc.) for which the organism's strategies
are appropriate and hence within which the species reaches a sufficiently high fitness to survive and fend off predators. 
Organisms may also have strategies to alter their environment so that some of their strategies become more possible or more optimal, 
a process called {\itshape niche construction} \citep{Odling-Smee:2003} \is{fitness landscape}

@@

The previous chapter already showed that there are 
different strategies for colour language: basic colour terms, graded membership (`very green'), compounds (`sky blue'), 
brightness qualification (`light green'), etc. All these strategies use a particular colour space, more specifically 
the CIE L*u*v* space where the L* dimension represents brightness (ranging from black to white), the u* 
dimension represents the red-green opponent channel, 
and the v* dimension the yellow-blue channel. This space was chosen 
because the distance between two colours in the L*u*v* space accurately represents 
the psychological distances between these colours as perceived by human subjects. \is{Color Description Game}

Let us focus on a family of strategies that all use a prototype-based approach, i.e. they perform categorisation using 
a standard nearest neighbour classification algorithm operating over an inventory of prototypical points, 
as also used in the Proper Naming Game and the Basic Colour Naming Game discussed in the previous chapter. 
Three variant strategies are imaginable (and can be found in human languages): \enlargethispage{2\baselineskip}
\begin{enumerate}
\item The {\itshape brightness-only strategy} which uses only the L* dimension (dark, grey, bright, etc.).
\item The {\itshape hue-only strategy} which uses only the hue dimension (blue, yellow, green, etc.). 
\item And the {\itshape full-colour strategy} which uses all dimensions. 
\end{enumerate}
In order to understand the niche of each strategy, Bleys has set up experiments where the context is controlled 
with respect to the importance of hue. 
For example, agents can be presented with contexts with samples where the hue is the same but brightness is different on the one 
hand versus contexts with samples where the brightness is the same but hue is different. In between we get a mixed 
importance of hue.

\begin{figure}[htb!]
\centerline{
\includegraphics[width=0.6\textwidth]{chap12/figs/selective-advantage-bar.pdf}
}
\caption{{Niche analysis for three different 
colour strategies: brightness-only, brightness-hue, and hue-only. 
The properties of the context vary on the x-axis and the y-axis shows performance of each strategy.
Only one niche dimension is examined, namely communicative success. 
We see that each strategy has its own niche for which it is optimal, but the full-colour strategy is globally 
performing the best across contexts, which may explain why it has become preferred in most language families. 
\label{fig:bar}}
} 
\end{figure}

Bleys initialised the three populations with the three different strategies and let them first learn an interpretation of 
the Spanish focal colours (with prototypes in brightness\-only, hue-only, and full-colour spaces). Then he measured 
performance for communicative success for the varying ecological conditions. 
Not unexpectedly, he found that depending on the percentage of the importance of hue differences in the context, a different strategy 
is more optimal (see \figref{fig:bar}). Not surprisingly, a hue-only strategy fails completely when the samples in each  
context have equal hue but different brightness, whereas a brightness-only strategy fails when the samples have 
equal brightness but different hue.

@@

The Bleys experiment concerns the brightness-based and the full-colour strategy. Both use 
a prototype-based nearest neighbour categorisation. We know from earlier experiments with the Proper Naming Game
and the Action Naming Game that this strategy enables a population of agents 
to self-organise a colour lexicon from scratch. \figref{fig:full-cs-dynamics} illustrates (on the left) how 
communicative success increases and the lexicon becomes more coherent, and (on the right) how the prototypes
of the different agents gradually expand and become similar to be almost identical after 1200 games per agent, when 
the system has stabilised to 6 basic colours.

@@

The {\itshape Brightness Prototype Strategy} is similar to the Basic Colour Strategy, but instead of taking all three dimensions into account, only the L* dimension of both the stimuli in the context and the prototypes of the colour categories are compared. While learning, the prototype of the used colour category is shifted on the L* axis towards the L* value of the topic. During invention, only the L* value of the topic is considered relevant. \figref{fig:brightness-dynamics} shows that this strategy is also adequate to allow a population of agents to self-organise and coordinate a colour lexicon from scratch. The resulting colour lexicon now consists of different shades of grey. The set of prototypes also expands and becomes similar across agents.

@@

Agents are initialised with a random value for the strategy-score for the Brightness-based versus
Full-colour based strategy. When the dynamics described here takes its 
course (which includes the shifting of prototypes by speakers and hearers to be maximally adapted to the 
contexts that are presented), we observe several possibilities. In one simulation run (see \figref{fig:strategies-one}), 
the brightness-based strategy becomes and remains dominant. It could have been the brightness 
or the full-colour strategy, depending on small fluctuations in the early stages. 
In another simulation run we see that one strategy becomes dominant first (in this case the brightness strategy) 
to be overtaken later by another strategy (i.c.\ the full-colour strategy), 
as attested in the history of English colour terms (see \figref{fig:strategies-dynamics}). 
The two strategies continue to co-exist. Brightness is still used in 
circumstances where this gives a higher chance of communicative success, for example when colour chips are 
close in hue but distinct in brightness or when there is a word which has most of its success in the brightness 
dimension.

@@

\begin{figure}
\centerline{\includegraphics[width=.9\linewidth]{chap11/figs/near-the-box2.pdf}}
\caption{Example of a conceptualisation for the phrase `near to the box'. {\itshape Top:} 
Leftmost: The situation as perceived by the robot. Middle: situation after a geometric transform where the box is 
made the central perspective. Right: Area around the box which reflects (in terms of darkness of the colour) how 
near objects are to the box. {\itshape Bottom:} IRL constraint network for this conceptualisation.
\label{fig:spatial-strategies}}
\end{figure}

@@

Once an adequate conceptualisation is found and it leads to a successful game, modular sections of the network can be 
abstracted and used as the basis of reusable conceptualisation strategies in the future. Each conceptualisation 
strategy has its own score which is updated using the lateral inhibition dynamics as used for 
the lexical or grammatical constructions as explained in earlier chapters. The frequency of use and success 
of a strategy depends partly on whether the ``niche'' in which a strategy is most effective occurs frequently, partly 
on what is preferred by the rest of the population, and other criteria such as complexity.

@@

Are there general meta-strategies that can be used to generate the kind of lexico\-gram\-mat\-ical strategies 
that can in turn generate all the language systems we observe 
in today's languages? This is a very profound question, related to the ongoing debate in linguistics on the existence of 
a universal language acquisition device. A lot of ink has been spilled but we need to do profound computational modelling 
of meta-strategies and thorough experiments to find out.

@@

For {\itshape lexical strategies}, the diagnostics consist in checking whether there is part 
of the meaning that is not yet covered by any lexical item. The repair then consists in introducing 
a (possibly new) word and introducing a new lexical construction for this word. 
This strategy is already used by the Talking Heads. When part of a meaning cannot be covered, a new word form is randomly generated
and used from then on for this meaning, but possibly in competition with other words that were later acquired for the same 
meaning.

\newpage
A study of ongoing change in the vocabularies of human speakers shows that the invention of a brand new word form happens seldom. 
Usually an existing wordform is recruited to express an uncovered meaning. The meaning of the already existing word form
expresses aspects of the novel meaning to be expressed but there is not an exact overlap. However, the existing word is sufficiently 
suggestive so that the hearer can guess what meaning is intended by the speaker. For example, the word `mouse' becomes 
adopted for computer mouse based on its appearance. Often a compound of several words is used, each of them suggesting 
other aspects of the meaning, such as `bitcoin', which suggests digital (`bit') and currency (`coin').  
There are several other strategies, for example, an abbreviation of a long word becomes itself a single new word, 
such as `MOOC', meaning `massively open on-line course' and now standing for a course made available through the internet. 
There are also playful variations created with existing words, as for example
`wackadoo' which means `bizarre person' and probably originally came from the word `wacky'.

@@

Historical linguists have researched many concrete examples of how agreement systems arise.\footnote{See for example: \cite{Lehmann:1988}, \cite{Luraghi:2011}.}
There are invariably the following phases: 
\begin{enumerate}
\item Agreement markers derive from reusing existing words, such as pronouns or classifiers. 
\item The markers derived from independent words become shorter, they lose part of their form, 
then become clitics and later affixes. For example, the Dyirbal (Australian aboriginal)
agreement marker {\itshape m-} evolved from the classifier {\itshape mayi}, which means 'non-flesh food'. 
\item The features that agreement markers express, which are initially semantically 
grounded, invariably become more abstract and get to be used in 
a purely conventional manner. For example, the masculine/feminine 
gender distinction has its basis in male/female sex, but is then arbitrarily applied to inanimate objects, so that table 
might be masculine in one language (German: {\itshape der Tisch}) but feminine in another (French: {\itshape la table}). 
\end{enumerate}
Starting from a game of reference, similar to the Talking Heads experiment, populations of agents were programmed 
first to build a ``formal'' agreement system, i.e. a system based on formal markers without any meaning, and then 
a meaningful grammatical system, with grammatical markers that express semantic features. The experiment was 
then extended with further strategies that optimise three niche factors:
\begin{enumerate}
\item The recruitment of existing words as markers makes the acquisition of new markers more accurate because it is based 
on existing words and so the meaning can be guessed more easily. This helps to propagate the markers faster in the population. 
\item The phonological reduction of word forms leads to less articulatory effort, as we have already seen with the German 
article system. 
\item Conventionalisation leads to a smaller inventory, and hence less memory requirements and better preservation of the 
language system. 
\end{enumerate}
